Internal models of type theory and internal higher categories for the Univalent Foundations

### 2014-03-02 article-models-tt.tex
small impro
using the new judgment macro

It should be noted that this is just a prototype description. It may be the case
that various other elements have to be added yet, such as terms
$\jterm{\ctxext{\Gamma}{\subst{x}{P}}}{\ctxwk{\subst{x}{P}}{\ctxext{A}{P}}}{i}$
including the fiber $\subst{x}{P}$ into the extended family $\ctxext{A}{P}$ over
$\Gamma$. Moreover, it hasn't been thoroughly tested whether the theory presented
here lives up to its promise.

@@

\begin{align}
& \inference{\jfam{\Gamma}{A}}{\jctx{\ctxext{\Gamma}{A}}}
& & \inference{\jctxeq{\Gamma}{\Delta}\qquad\jfameq{\Gamma}{A}{B}}{\jctxeq{\ctxext{\Gamma}{A}}{\ctxext{\Delta}{B}}}\\
& \inference{\jfam{\ctxext{\Gamma}{A}}{P}}{\jfam{\Gamma}{\ctxext{A}{P}}}
& & \inference{\jfameq{\Gamma}{A}{B}\qquad\jfameq{\ctxext{\Gamma}{A}}{P}{Q}}{\jfameq{\Gamma}{\ctxext{A}{P}}{\ctxext{B}{Q}}}
\end{align}

@@

\begin{align}
& \inference{\jfam{\Gamma}{A}\qquad\jfam{\ctxext{\Gamma}{A}}{P}}
  {\jctxeq{\ctxext{{\Gamma}{A}}{P}}{\ctxext{\Gamma}{{A}{P}}}}\\
& \inference{\jfam{\ctxext{\Gamma}{A}}{P}\qquad\jfam{\ctxext{{\Gamma}{A}}{P}}{Q}}
  {\jfameq{\Gamma}{\ctxext{{A}{P}}{Q}}{\ctxext{A}{{P}{Q}}}}
\end{align}

@@

 In the following inference rules we assume that $\jfam{\Gamma}{A}$ and in the
 rules asserting a judgmental equality we assume furthermore that 
 $\jfameq{\Gamma}{A}{A'}$.
\begin{align}
& \inference{\jfam{\Gamma}{B}}{\jfam{\ctxext{\Gamma}{A}}{\ctxwk{A}{B}}}
& & \inference{\jfameq{\Gamma}{B}{B'}}{\jfameq{\ctxext{\Gamma}{A}}{\ctxwk{A}{B}}{\ctxwk{A'}{B'}}}\\
& \inference{\jfam{\ctxext{\Gamma}{B}}{Q}}
{\jfam{\ctxext{{\Gamma}{A}}{\ctxwk{A}{B}}}{\ctxwk{A}{Q}}}
& & \inference{\jfameq{\ctxext{\Gamma}{B}}{Q}{Q'}}
{\jfameq{\ctxext{{\Gamma}{A}}{\ctxwk{A}{B}}}{\ctxwk{A}{Q}}{\ctxwk{A'}{Q'}}}\\
& \inference{\jterm{\Gamma}{B}{y}}{\jterm{\ctxext{\Gamma}{A}}{\ctxwk{A}{B}}{\ctxwk{A}{y}}}
& & \inference{\jtermeq{\Gamma}{B}{y}{y'}}{\jtermeq{\ctxext{\Gamma}{A}}{\ctxwk{A}{B}}{\ctxwk{A}{y}}{\ctxwk{A'}{y'}}}\\
& \inference{\jterm{\ctxext{\Gamma}{B}}{Q}{g}}{\jterm{\ctxext{{\Gamma}{A}}{\ctxwk{A}{B}}}{\ctxwk{A}{Q}}{\ctxwk{A}{g}}}
& & \inference{\jtermeq{\ctxext{\Gamma}{B}}{Q}{g}{g'}}
{\jtermeq{\ctxext{{\Gamma}{A}}{\ctxwk{A}{B}}}{\ctxwk{A}{Q}}{\ctxwk{A}{g}}{\ctxwk{A'}{g'}}}
\end{align}

@@

\subsubsection{Weakening is compatible with the empty context}
The following rules express that when the empty context or context family is
weakened, the result is the empty context family.
\begin{align}
& \inference{\jctx{\Gamma}}{\jfameq{\Gamma}{\ctxwk{\Gamma}{\emptyc}}{\emptyf}}\\
& \inference{\jfam{\Gamma}{A}}{\jfameq{\ctxext{\Gamma}{A}}{\ctxwk{A}{\emptyf}}{\emptyf}}
\end{align}
Weakening by the empty family $\emptyf$ over a context $\Gamma$ leaves families, 
their terms, families over those families and
terms of those unchanged:
\begin{align}
& \inference{\jfam{\Gamma}{B}}{\jfameq{\Gamma}{\ctxwk{\emptyf}{B}}{B}}\\
& \inference{\jterm{\Gamma}{B}{y}}{\jtermeq{\Gamma}{B}{\ctxwk{\emptyf}{y}}{y}}\\
& \inference{\jfam{\ctxext{\Gamma}{B}}{Q}}{\jfameq{\ctxext{\Gamma}{B}}{\ctxwk{\emptyf}{Q}}{Q}}\\
& \inference{\jterm{\ctxext{\Gamma}{B}}{Q}{g}}{\jtermeq{\ctxext{\Gamma}{B}}{Q}{\ctxwk{\emptyf}{g}}{g}}
\end{align}

@@

In the following inference rules we assume that $\jfam{\Gamma}{A}$.
\begin{align}
& \inference
  {\jfam{\ctxext{{\Gamma}{B}}{Q}}{R}}
  {\jfameq
    {\ctxext{{\Gamma}{A}}{\ctxwk{A}{B}}}
    {\ctxwk{A}{\ctxext{Q}{R}}}
    {\ctxext{\ctxwk{A}{Q}}{\ctxwk{A}{R}}}}
\end{align}

\subsubsection{Weakening is compatible with itself}\label{comp-ww}
We state judgmental equality rules expressing
that weakening is compatible with itself. These rules state that the following
diagram commutes given any two families $A$ and $B$ in context $\Gamma$:
\begin{equation*}
\begin{tikzcd}[column sep=huge]
\jfam{\Gamma}{\blank} \ar{r}{C\mapsto\ctxwk{B}{C}} \ar{d}[swap]{C\mapsto\ctxwk{A}{C}} & \jfam{\ctxext{\Gamma}{B}}{\blank} \ar{d}{Q\mapsto\ctxwk{A}{Q}}\\
\jfam{\ctxext{\Gamma}{A}}{\blank} \ar{r}[swap]{P\mapsto\ctxwk{{A}{B}}{P}} & \jfam{\ctxext{{\Gamma}{A}}{\ctxwk{A}{B}}}{\blank}
\end{tikzcd}
\end{equation*}
Thus, we get the following set of inference rules:
\begin{align}
& \inference{\jfam{\Gamma}{A}\qquad\jfam{\Gamma}{B}\qquad\jfam{\ctxext{\Gamma}{C}}{R}}
          {\jfameq{\ctxext{{{\Gamma}{A}}{\ctxwk{A}{B}}}{\ctxwk{{A}{B}}{{A}{C}}}}
            {\ctxwk{A}{{B}{R}}}
            {\ctxwk{{A}{B}}{{A}{R}}}}\label{comp-ww-f}\\
& \inference{\jfam{\Gamma}{A}\qquad\jfam{\Gamma}{B}\qquad\jterm{\ctxext{\Gamma}{C}}{R}{t}}
          {\jtermeq{\ctxext{{{\Gamma}{A}}{\ctxwk{A}{B}}}{\ctxwk{{A}{B}}{{A}{C}}}}
            {\ctxwk{{A}{B}}{{A}{R}}}{\ctxwk{A}{{B}{t}}}{\ctxwk{{A}{B}}{{A}{t}}}}
            \label{comp-ww-t}
\end{align}

@@

In the following inference rules we assume that
$\jfam{\Gamma}{A}$ and $\jfam{\ctxext{\Gamma}{A}}{P}$. 
\begin{align}
& \inference{\jfam{\ctxext{\Gamma}{B}}{Q}}
    {\jfameq{\ctxext{{{\Gamma}{A}}{P}}{\ctxwk{P}{{A}{B}}}}{\ctxwk{\ctxext{A}{P}}{Q}}{\ctxwk{P}{{A}{Q}}}}\label{comp-ew-f}\\
& \inference{\jterm{\ctxext{\Gamma}{B}}{Q}{g}}
    {\jtermeq{\ctxext{{{\Gamma}{A}}{P}}{\ctxwk{P}{{A}{B}}}}{\ctxwk{P}{{A}{Q}}}{\ctxwk{\ctxext{A}{P}}{g}}{\ctxwk{P}{{A}{g}}}}\label{comp-ew-t}
\end{align}

@@

\begin{align}
& \inference{\jfam{\ctxext{\Gamma}{A}}{P}}{\jfam{\Gamma}{\subst{x}{P}}}
& & \inference{\jfameq{\ctxext{\Gamma}{A}}{P}{P'}}{\jfameq{\Gamma}{\subst{x}{P}}{\subst{x'}{P'}}}\\
& \inference{\jfam{\ctxext{{\Gamma}{A}}{P}}{Q}}{\jfam{\ctxext{\Gamma}{\subst{x}{P}}}{\subst{x}{Q}}}
& & \inference{\jfameq{\ctxext{{\Gamma}{A}}{P}}{Q}{Q'}}{\jfameq{\ctxext{\Gamma}{\subst{x}{P}}}{\subst{x}{Q}}{\subst{x'}{Q'}}}\\
& \inference{\jterm{\ctxext{\Gamma}{A}}{P}{f}}{\jterm{\Gamma}{\subst{x}{P}}{\subst{x}{f}}}
& & \inference{\jtermeq{\ctxext{\Gamma}{A}}{P}{f}{f'}}{\jtermeq{\Gamma}{\subst{x}{P}}{\subst{x}{f}}{\subst{x'}{f'}}}\\
& \inference{\jterm{\ctxext{{\Gamma}{A}}{P}}{Q}{g}}{\jterm{\ctxext{\Gamma}{\subst{x}{P}}}{\subst{x}{Q}}{\subst{x}{g}}}
& &\inference{\jtermeq{\ctxext{{\Gamma}{A}}{P}}{Q}{g}{g'}}{\jtermeq{\ctxext{\Gamma}{\subst{x}{P}}}{\subst{x}{Q}}{\subst{x}{g}}{\subst{x'}{g'}}}
\end{align}

@@

\subsubsection{Substitution is compatible with the empty context}
The fibers of the empty family are the empty families:
\begin{align}
& \inference{\jterm{\Gamma}{A}{x}}{\jfameq{\Gamma}{\subst{x}{\emptyf}}{\emptyf}}\\
& \inference{\jterm{\Gamma}{A}{x}\quad\jfam{\ctxext{\Gamma}{A}}{P}}{\jfameq{\ctxext{\Gamma}{\subst{x}{P}}}{\subst{x}{\emptyf}}{\emptyf}}
\end{align}

The following rules assert that substituting by the term $\jterm{\Gamma}{\emptyf}{\emptytm}$
leaves everything unchanged.
\begin{align}
& \inference{\jfam{\Gamma}{A}}{\jfameq{\Gamma}{\subst{\emptytm}{A}}{A}}\\
& \inference{\jterm{\Gamma}{A}{x}}{\jtermeq{\Gamma}{A}{\subst{\emptytm}{x}}{x}}\\
& \inference{\jfam{\ctxext{\Gamma}{A}}{P}}{\jfameq{\ctxext{\Gamma}{A}}{\subst{\emptytm}{P}}{P}}\\
& \inference{\jterm{\ctxext{\Gamma}{A}}{P}{f}}{\jtermeq{\ctxext{\Gamma}{A}}{P}{\subst{\emptytm}{f}}{f}}.
\end{align}

\subsubsection{Substitution is compatible with extension}\label{comp-se}
Suppose $\jterm{\Gamma}{A}{x}$ in all of the following inference rule.
\begin{align}
& \inference{\jfam{\ctxext{{{\Gamma}{A}}{P}}{Q}}{R}}
  {\jfameq{\ctxext{\Gamma}{\subst{x}{P}}}{\subst{x}{\ctxext{Q}{R}}}{\ctxext{\subst{x}{Q}}{\subst{x}{R}}}}
\end{align}

\subsubsection{Substitution is compatible with weakening}\label{comp-sw}
The rules asserting the compatibility of substitution with weakening assert
that the following diagram commutes for any $\jterm{\Gamma}{A}{x}$ and any
$\jfam{\ctxext{\Gamma}{A}}{P}$:
\begin{equation*}
\begin{tikzcd}[column sep=huge]
\jfam{\ctxext{\Gamma}{A}}{\blank} \ar{d}[swap]{Q\mapsto\subst{x}{Q}} \ar{r}{Q\mapsto\ctxwk{P}{Q}} & \jfam{\ctxext{{\Gamma}{A}}{P}}{\blank} \ar{d}{R\mapsto\subst{x}{R}}\\ 
\jfam{\Gamma}{\blank} \ar{r}[swap]{B\mapsto\ctxwk{\subst{x}{P}}{B}} & \jfam{\ctxext{\Gamma}{\subst{x}{P}}}{\blank}
\end{tikzcd}
\end{equation*}
We plug in an extra layer of families to cover the most general case at once.
In the following inference rules we assume that $\jterm{\Gamma}{A}{x}$ and
$\jfam{\ctxext{\Gamma}{A}}{P}$:
\begin{align}
& \inference
  {\jfam{\ctxext{{\Gamma}{A}}{Q}}{R}}
  {\jfameq{\ctxext{{\Gamma}{\subst{x}{P}}}{\subst{x}{\ctxwk{P}{Q}}}}{\subst{x}{\ctxwk{P}{R}}}{\ctxwk{\subst{x}{P}}{\subst{x}{R}}}}\label{comp-sw-f}\\
& \inference
    {\jterm{\ctxext{{\Gamma}{A}}{Q}}{R}{h}}
    {\jtermeq
      {\ctxext{{\Gamma}{\subst{x}{P}}}{\subst{x}{\ctxwk{P}{Q}}}}
      {\subst{x}{\ctxwk{P}{R}}}
      {\subst{x}{\ctxwk{P}{h}}}
      {\ctxwk{\subst{x}{P}}{\subst{x}{h}}}}\label{comp-sw-t}
\end{align}

@@

In the following inference rules we assume
$\jterm{\Gamma}{A}{x}$ and $\jterm{\ctxext{\Gamma}{A}}{P}{f}$.

\begin{align}
&\inference{\jfam{\ctxext{{{\Gamma}{A}}{P}}{Q}}{R}}
{\jfameq{\ctxext{\Gamma}{\subst{x}{{f}{Q}}}}{\subst{x}{{f}{R}}}{\subst{{x}{f}}{{x}{R}}}}\label{comp-ss-f}\\
&\inference{\jterm{\ctxext{{{\Gamma}{A}}{P}}{Q}}{R}{h}}
{\jtermeq{\ctxext{\Gamma}{\subst{x}{{f}{Q}}}}{\subst{x}{{f}{R}}}{\subst{x}{{f}{h}}}{\subst{{x}{f}}{{x}{h}}}}\label{comp-ss-t}
\end{align}

@@

\begin{align}
& \inference{\jfam{\ctxext{{\Gamma}{B}}{Q}}{R}}{\jfameq{\ctxext{{\Gamma}{A}}{\ctxwk{A}{\subst{y}{Q}}}}{\ctxwk{A}{\subst{y}{R}}}{\subst{\ctxwk{A}{y}}{\ctxwk{A}{R}}}}\label{comp-ws-f}\\
& \inference{\jterm{\ctxext{{\Gamma}{B}}{Q}}{R}{h}}{\jtermeq{\ctxext{{\Gamma}{A}}{\ctxwk{A}{\subst{y}{Q}}}}{\ctxwk{A}{\subst{y}{R}}}{\ctxwk{A}{\subst{y}{h}}}{\subst{\ctxwk{A}{y}}{\ctxwk{A}{h}}}}\label{comp-ws-t}
\end{align}

@@

\begin{align}
& \inference{\jfam{\Gamma}{A}\qquad\jfam{\ctxext{\Gamma}{B}}{Q}\qquad\jterm{\Gamma}{A}{x}}{\jfameq{\ctxext{\Gamma}{B}}{\subst{x}{\ctxwk{A}{Q}}}{Q}}\label{defn-ws-3}\\
& \inference{\jterm{\Gamma}{A}{x}\qquad\jterm{\ctxext{\Gamma}{B}}{Q}{g}}{\jtermeq{\ctxext{\Gamma}{B}}{Q}{\subst{x}{\ctxwk{A}{g}}}{g}}\label{defn-ws-4}
\end{align}

Using the rules of the compatibility of substitution with weakening and of the
compatibility of weakening with itself, we see that we can show
\begin{equation*}
\jfameq{\ctxext{\Gamma}{A}}{\subst{f}{\ctxwk{A}{{B}{C}}}}{\ctxwk{A}{C}}
\end{equation*}
for any three families $A$, $B$ and $C$ in context $\Gamma$ and any $\jterm{\ctxext{\Gamma}{A}}{\ctxwk{A}{B}}{f}$.
It follows that for $\jterm{\ctxext{\Gamma}{B}}{\ctxwk{B}{C}}{g}$ we have
\begin{equation*}
\jterm{\ctxext{\Gamma}{A}}{\ctxwk{A}{C}}{\subst{f}{\ctxwk{A}{g}}}
\end{equation*}
The term $\jcomp{A}{f}{g}$ is the composition of $g$ with $f$. We may call
a term $\jterm{\ctxext{\Gamma}{A}}{\ctxwk{A}{B}}{f}$ a morphism from $A$ to $B$.

>>>>

It should be noted that this is just a prototype description. It may be the case
that various other elements have to be added yet, such as terms
$\jterm{{\Gamma}{\subst{x}{P}}}{\ctxwk{\subst{x}{P}}{\ctxext{A}{P}}}{i}$
including the fiber $\subst{x}{P}$ into the extended family $\ctxext{A}{P}$ over
$\Gamma$. Moreover, it hasn't been thoroughly tested whether the theory presented
here lives up to its promise.

@@

\begin{align}
& \inference{\jfam{\Gamma}{A}}{\jctx{\ctxext{\Gamma}{A}}}
& & \inference{\jctxeq{\Gamma}{\Delta}\qquad\jfameq{\Gamma}{A}{B}}{\jctxeq{\ctxext{\Gamma}{A}}{\ctxext{\Delta}{B}}}\\
& \inference{\jfam{{\Gamma}{A}}{P}}{\jfam{\Gamma}{\ctxext{A}{P}}}
& & \inference{\jfameq{\Gamma}{A}{B}\qquad\jfameq{{\Gamma}{A}}{P}{Q}}{\jfameq{\Gamma}{\ctxext{A}{P}}{\ctxext{B}{Q}}}
\end{align}

@@

\begin{align}
& \inference{\jfam{\Gamma}{A}\qquad\jfam{{\Gamma}{A}}{P}}
  {\jctxeq{\ctxext{{\Gamma}{A}}{P}}{\ctxext{\Gamma}{{A}{P}}}}\\
& \inference{\jfam{{\Gamma}{A}}{P}\qquad\jfam{{{\Gamma}{A}}{P}}{Q}}
  {\jfameq{\Gamma}{\ctxext{{A}{P}}{Q}}{\ctxext{A}{{P}{Q}}}}
\end{align}

@@

 In the following inference rules we assume that $\jfam{\Gamma}{A}$ and in the
 rules asserting a judgmental equality we assume furthermore that 
 $\jfameq{\Gamma}{A}{A'}$.
\begin{align}
& \inference{\jfam{\Gamma}{B}}{\jfam{{\Gamma}{A}}{\ctxwk{A}{B}}}
& & \inference{\jfameq{\Gamma}{B}{B'}}{\jfameq{{\Gamma}{A}}{\ctxwk{A}{B}}{\ctxwk{A'}{B'}}}\\
& \inference{\jfam{{\Gamma}{B}}{Q}}
{\jfam{{{\Gamma}{A}}{\ctxwk{A}{B}}}{\ctxwk{A}{Q}}}
& & \inference{\jfameq{{\Gamma}{B}}{Q}{Q'}}
{\jfameq{{{\Gamma}{A}}{\ctxwk{A}{B}}}{\ctxwk{A}{Q}}{\ctxwk{A'}{Q'}}}\\
& \inference{\jterm{\Gamma}{B}{y}}{\jterm{{\Gamma}{A}}{\ctxwk{A}{B}}{\ctxwk{A}{y}}}
& & \inference{\jtermeq{\Gamma}{B}{y}{y'}}{\jtermeq{{\Gamma}{A}}{\ctxwk{A}{B}}{\ctxwk{A}{y}}{\ctxwk{A'}{y'}}}\\
& \inference{\jterm{{\Gamma}{B}}{Q}{g}}{\jterm{{{\Gamma}{A}}{\ctxwk{A}{B}}}{\ctxwk{A}{Q}}{\ctxwk{A}{g}}}
& & \inference{\jtermeq{{\Gamma}{B}}{Q}{g}{g'}}
{\jtermeq{{{\Gamma}{A}}{\ctxwk{A}{B}}}{\ctxwk{A}{Q}}{\ctxwk{A}{g}}{\ctxwk{A'}{g'}}}
\end{align}

@@

\subsubsection{Weakening is compatible with the empty context}
The following rules express that when the empty context or context family is
weakened, the result is the empty context family.
\begin{align}
& \inference{\jctx{\Gamma}}{\jfameq{\Gamma}{\ctxwk{\Gamma}{\emptyc}}{\emptyf}}\\
& \inference{\jfam{\Gamma}{A}}{\jfameq{{\Gamma}{A}}{\ctxwk{A}{\emptyf}}{\emptyf}}
\end{align}
Weakening by the empty family $\emptyf$ over a context $\Gamma$ leaves families, 
their terms, families over those families and
terms of those unchanged:
\begin{align}
& \inference{\jfam{\Gamma}{B}}{\jfameq{\Gamma}{\ctxwk{\emptyf}{B}}{B}}\\
& \inference{\jterm{\Gamma}{B}{y}}{\jtermeq{\Gamma}{B}{\ctxwk{\emptyf}{y}}{y}}\\
& \inference{\jfam{{\Gamma}{B}}{Q}}{\jfameq{{\Gamma}{B}}{\ctxwk{\emptyf}{Q}}{Q}}\\
& \inference{\jterm{{\Gamma}{B}}{Q}{g}}{\jtermeq{{\Gamma}{B}}{Q}{\ctxwk{\emptyf}{g}}{g}}
\end{align}

@@

In the following inference rules we assume that $\jfam{\Gamma}{A}$.
\begin{align}
& \inference
  {\jfam{{{\Gamma}{B}}{Q}}{R}}
  {\jfameq
    {\ctxext{{\Gamma}{A}}{\ctxwk{A}{B}}}
    {\ctxwk{A}{\ctxext{Q}{R}}}
    {\ctxext{\ctxwk{A}{Q}}{\ctxwk{A}{R}}}}
\end{align}

\subsubsection{Weakening is compatible with itself}\label{comp-ww}
We state judgmental equality rules expressing
that weakening is compatible with itself. These rules state that the following
diagram commutes given any two families $A$ and $B$ in context $\Gamma$:
\begin{equation*}
\begin{tikzcd}[column sep=huge]
\jfam{\Gamma}{\blank} \ar{r}{C\mapsto\ctxwk{B}{C}} \ar{d}[swap]{C\mapsto\ctxwk{A}{C}} & \jfam{{\Gamma}{B}}{\blank} \ar{d}{Q\mapsto\ctxwk{A}{Q}}\\
\jfam{{\Gamma}{A}}{\blank} \ar{r}[swap]{P\mapsto\ctxwk{{A}{B}}{P}} & \jfam{{{\Gamma}{A}}{\ctxwk{A}{B}}}{\blank}
\end{tikzcd}
\end{equation*}
Thus, we get the following set of inference rules:
\begin{align}
& \inference{\jfam{\Gamma}{A}\qquad\jfam{\Gamma}{B}\qquad\jfam{{\Gamma}{C}}{R}}
          {\jfameq{{{{\Gamma}{A}}{\ctxwk{A}{B}}}{\ctxwk{{A}{B}}{{A}{C}}}}
            {\ctxwk{A}{{B}{R}}}
            {\ctxwk{{A}{B}}{{A}{R}}}}\label{comp-ww-f}\\
& \inference{\jfam{\Gamma}{A}\qquad\jfam{\Gamma}{B}\qquad\jterm{{\Gamma}{C}}{R}{t}}
          {\jtermeq{{{{\Gamma}{A}}{\ctxwk{A}{B}}}{\ctxwk{{A}{B}}{{A}{C}}}}
            {\ctxwk{{A}{B}}{{A}{R}}}{\ctxwk{A}{{B}{t}}}{\ctxwk{{A}{B}}{{A}{t}}}}
            \label{comp-ww-t}
\end{align}

@@

In the following inference rules we assume that
$\jfam{\Gamma}{A}$ and $\jfam{{\Gamma}{A}}{P}$. 
\begin{align}
& \inference{\jfam{{\Gamma}{B}}{Q}}
    {\jfameq{{{{\Gamma}{A}}{P}}{\ctxwk{P}{{A}{B}}}}{\ctxwk{\ctxext{A}{P}}{Q}}{\ctxwk{P}{{A}{Q}}}}\label{comp-ew-f}\\
& \inference{\jterm{{\Gamma}{B}}{Q}{g}}
    {\jtermeq{{{{\Gamma}{A}}{P}}{\ctxwk{P}{{A}{B}}}}{\ctxwk{P}{{A}{Q}}}{\ctxwk{\ctxext{A}{P}}{g}}{\ctxwk{P}{{A}{g}}}}\label{comp-ew-t}
\end{align}

@@

\begin{align}
& \inference{\jfam{{\Gamma}{A}}{P}}{\jfam{\Gamma}{\subst{x}{P}}}
& & \inference{\jfameq{{\Gamma}{A}}{P}{P'}}{\jfameq{\Gamma}{\subst{x}{P}}{\subst{x'}{P'}}}\\
& \inference{\jfam{{{\Gamma}{A}}{P}}{Q}}{\jfam{{\Gamma}{\subst{x}{P}}}{\subst{x}{Q}}}
& & \inference{\jfameq{{{\Gamma}{A}}{P}}{Q}{Q'}}{\jfameq{{\Gamma}{\subst{x}{P}}}{\subst{x}{Q}}{\subst{x'}{Q'}}}\\
& \inference{\jterm{{\Gamma}{A}}{P}{f}}{\jterm{\Gamma}{\subst{x}{P}}{\subst{x}{f}}}
& & \inference{\jtermeq{{\Gamma}{A}}{P}{f}{f'}}{\jtermeq{\Gamma}{\subst{x}{P}}{\subst{x}{f}}{\subst{x'}{f'}}}\\
& \inference{\jterm{{{\Gamma}{A}}{P}}{Q}{g}}{\jterm{{\Gamma}{\subst{x}{P}}}{\subst{x}{Q}}{\subst{x}{g}}}
& &\inference{\jtermeq{{{\Gamma}{A}}{P}}{Q}{g}{g'}}{\jtermeq{{\Gamma}{\subst{x}{P}}}{\subst{x}{Q}}{\subst{x}{g}}{\subst{x'}{g'}}}
\end{align}

@@

\subsubsection{Substitution is compatible with the empty context}
The fibers of the empty family are the empty families:
\begin{align}
& \inference{\jterm{\Gamma}{A}{x}}{\jfameq{\Gamma}{\subst{x}{\emptyf}}{\emptyf}}\\
& \inference{\jterm{\Gamma}{A}{x}\quad\jfam{{\Gamma}{A}}{P}}{\jfameq{{\Gamma}{\subst{x}{P}}}{\subst{x}{\emptyf}}{\emptyf}}
\end{align}

The following rules assert that substituting by the term $\jterm{\Gamma}{\emptyf}{\emptytm}$
leaves everything unchanged.
\begin{align}
& \inference{\jfam{\Gamma}{A}}{\jfameq{\Gamma}{\subst{\emptytm}{A}}{A}}\\
& \inference{\jterm{\Gamma}{A}{x}}{\jtermeq{\Gamma}{A}{\subst{\emptytm}{x}}{x}}\\
& \inference{\jfam{{\Gamma}{A}}{P}}{\jfameq{{\Gamma}{A}}{\subst{\emptytm}{P}}{P}}\\
& \inference{\jterm{{\Gamma}{A}}{P}{f}}{\jtermeq{{\Gamma}{A}}{P}{\subst{\emptytm}{f}}{f}}.
\end{align}

\subsubsection{Substitution is compatible with extension}\label{comp-se}
Suppose $\jterm{\Gamma}{A}{x}$ in all of the following inference rule.
\begin{align}
& \inference{\jfam{{{{\Gamma}{A}}{P}}{Q}}{R}}
  {\jfameq{{\Gamma}{\subst{x}{P}}}{\subst{x}{\ctxext{Q}{R}}}{\ctxext{\subst{x}{Q}}{\subst{x}{R}}}}
\end{align}

\subsubsection{Substitution is compatible with weakening}\label{comp-sw}
The rules asserting the compatibility of substitution with weakening assert
that the following diagram commutes for any $\jterm{\Gamma}{A}{x}$ and any
$\jfam{{\Gamma}{A}}{P}$:
\begin{equation*}
\begin{tikzcd}[column sep=huge]
\jfam{{\Gamma}{A}}{\blank} \ar{d}[swap]{Q\mapsto\subst{x}{Q}} \ar{r}{Q\mapsto\ctxwk{P}{Q}} & \jfam{{{\Gamma}{A}}{P}}{\blank} \ar{d}{R\mapsto\subst{x}{R}}\\ 
\jfam{\Gamma}{\blank} \ar{r}[swap]{B\mapsto\ctxwk{\subst{x}{P}}{B}} & \jfam{{\Gamma}{\subst{x}{P}}}{\blank}
\end{tikzcd}
\end{equation*}
We plug in an extra layer of families to cover the most general case at once.
In the following inference rules we assume that $\jterm{\Gamma}{A}{x}$ and
$\jfam{{\Gamma}{A}}{P}$:
\begin{align}
& \inference
  {\jfam{{{\Gamma}{A}}{Q}}{R}}
  {\jfameq{{{\Gamma}{\subst{x}{P}}}{\subst{x}{\ctxwk{P}{Q}}}}{\subst{x}{\ctxwk{P}{R}}}{\ctxwk{\subst{x}{P}}{\subst{x}{R}}}}\label{comp-sw-f}\\
& \inference
    {\jterm{{{\Gamma}{A}}{Q}}{R}{h}}
    {\jtermeq
      {\ctxext{{\Gamma}{\subst{x}{P}}}{\subst{x}{\ctxwk{P}{Q}}}}
      {\subst{x}{\ctxwk{P}{R}}}
      {\subst{x}{\ctxwk{P}{h}}}
      {\ctxwk{\subst{x}{P}}{\subst{x}{h}}}}\label{comp-sw-t}
\end{align}

@@

In the following inference rules we assume
$\jterm{\Gamma}{A}{x}$ and $\jterm{{\Gamma}{A}}{P}{f}$.

\begin{align}
&\inference{\jfam{{{{\Gamma}{A}}{P}}{Q}}{R}}
{\jfameq{{\Gamma}{\subst{x}{{f}{Q}}}}{\subst{x}{{f}{R}}}{\subst{{x}{f}}{{x}{R}}}}\label{comp-ss-f}\\
&\inference{\jterm{{{{\Gamma}{A}}{P}}{Q}}{R}{h}}
{\jtermeq{{\Gamma}{\subst{x}{{f}{Q}}}}{\subst{x}{{f}{R}}}{\subst{x}{{f}{h}}}{\subst{{x}{f}}{{x}{h}}}}\label{comp-ss-t}
\end{align}

@@

\begin{align}
& \inference{\jfam{{{\Gamma}{B}}{Q}}{R}}{\jfameq{{{\Gamma}{A}}{\ctxwk{A}{\subst{y}{Q}}}}{\ctxwk{A}{\subst{y}{R}}}{\subst{\ctxwk{A}{y}}{\ctxwk{A}{R}}}}\label{comp-ws-f}\\
& \inference{\jterm{{{\Gamma}{B}}{Q}}{R}{h}}{\jtermeq{{{\Gamma}{A}}{\ctxwk{A}{\subst{y}{Q}}}}{\ctxwk{A}{\subst{y}{R}}}{\ctxwk{A}{\subst{y}{h}}}{\subst{\ctxwk{A}{y}}{\ctxwk{A}{h}}}}\label{comp-ws-t}
\end{align}

@@

\begin{align}
& \inference{\jfam{\Gamma}{A}\qquad\jfam{{\Gamma}{B}}{Q}\qquad\jterm{\Gamma}{A}{x}}{\jfameq{{\Gamma}{B}}{\subst{x}{\ctxwk{A}{Q}}}{Q}}\label{defn-ws-3}\\
& \inference{\jterm{\Gamma}{A}{x}\qquad\jterm{{\Gamma}{B}}{Q}{g}}{\jtermeq{{\Gamma}{B}}{Q}{\subst{x}{\ctxwk{A}{g}}}{g}}\label{defn-ws-4}
\end{align}

Using the rules of the compatibility of substitution with weakening and of the
compatibility of weakening with itself, we see that we can show
\begin{equation*}
\jfameq{{\Gamma}{A}}{\subst{f}{\ctxwk{A}{{B}{C}}}}{\ctxwk{A}{C}}
\end{equation*}
for any three families $A$, $B$ and $C$ in context $\Gamma$ and any $\jterm{{\Gamma}{A}}{\ctxwk{A}{B}}{f}$.
It follows that for $\jterm{{\Gamma}{B}}{\ctxwk{B}{C}}{g}$ we have
\begin{equation*}
\jterm{{\Gamma}{A}}{\ctxwk{A}{C}}{\subst{f}{\ctxwk{A}{g}}}
\end{equation*}
The term $\jcomp{A}{f}{g}$ is the composition of $g$ with $f$. We may call
a term $\jterm{{\Gamma}{A}}{\ctxwk{A}{B}}{f}$ a morphism from $A$ to $B$.
